---
title: "Using CreateMyTaskSurv with the Lung Cancer Dataset"
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Using CreateMyTaskSurv with Lung Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

This vignette demonstrates how to use the `CreateMyTaskSurv` function from the
`mlr3SurvUtils` package to create survival tasks for the `mlr3` framework using the `lung` dataset from the `survival` package. The `lung` dataset contains survival data from patients with advanced lung cancer, including survival time, censoring status, and various covariates. We'll start with a minimal example for right-censored survival analysis (`type = "right"`) and then build on it by adding case-cohort study configurations and exploring a multi-state scenario. For simplicity, we'll use only complete cases from the dataset.

## Setup

First, load the required packages and the `lung` dataset:

```{r setup, message=FALSE}
library(mlr3)
library(mlr3proba)
library(data.table)
library(survival)
library(mlr3SurvUtils)
set.seed(123)  # For reproducibility
```

The `lung` dataset uses status with 1 = censored and 2 = dead, which differs from the usual 0/1 convention. 
We'll adjust this to match `CreateMyTaskSurv` expectations (0 = censored, 1 = event) and filter to complete cases


```{r data-prep}
lung_dt <- as.data.table(lung)
lung_dt <- lung_dt[complete.cases(lung_dt), ]  # Remove rows with NA
lung_dt[, status := status - 1]  # Convert to 0 = censored, 1 = dead
lung_dt[, id := .I]  # Add a unique identifier
```

This leaves us with `r nrow(lung_dt)` complete cases.

## Minimal Example: Right-Censored Survival Task

Let's create a basic survival task using `time` (survival time in days) and `status` (0/1) with `age` as a predictor:

```{r minimal-example}
target_info <- c(id = "lung_basic", time = "time", event = "status", type = "right")
backend_info <- list(
  id = "lung_data",
  primary_key = "id",
  feature_cols = c("age")
)

task_basic <- CreateMyTaskSurv(lung_dt, target_info, backend_info)
print(task_basic)
```

This task:

* Uses `type = "right"` for right-censored survival analysis.
* Has `age` as the only feature.
* Labels the task as "Dataset: lung_data; Task type: right".

You can inspect the data:

```{r inspect-basic}
head(task_basic$data())
```

## Adding Case-Cohort Study Configuration

Now, let's simulate a case-cohort study by randomly selecting a subcohort and applying 
weights using the `cch` argument. We'll assume 20% of the cohort is sampled as the subcohort:


```{r case-cohort}
# Simulate subcohort (20% random sample)
lung_dt[, subcoh := 0]
subcoh_size <- floor(nrow(lung_dt) * 0.2)
subcoh_ids <- sample(lung_dt$id, subcoh_size)
lung_dt[id %in% subcoh_ids, subcoh := 1]

# Define case-cohort configuration
cch_config <- list(
  subcohort = "subcoh",
  cch_weight = "Borgan1",
  sampling_fraction = 0.2,
  subcohort_only = FALSE
)

# Update target and backend info
target_info_cch <- c(id = "lung_cch", time = "time", event = "status", type = "right")
backend_info_cch <- list(
  id = "lung_data",
  primary_key = "id",
  feature_cols = c("age", "sex")
)

# Create task with case-cohort weights
task_cch <- CreateMyTaskSurv(
  lung_dt, 
  target_info_cch, 
  backend_info_cch, 
  cch = cch_config
)
print(task_cch)
```

Key points:

* `subcoh` indicates subcohort membership (1 = yes, 0 = no).
* `cch_weight` = "Borgan1" applies weights: subcohort censored cases get 1/0.2 = 5, events get 1, and non-subcohort censored cases are excluded (NA weights filtered out).
*  Added sex as an additional feature.
*  The task label reflects the case-cohort setup: "Dataset: lung_data; Task type: right; CCH weights: Borgan1; Sampling fraction: 0.2".

Inspect the data with weights:

```{r inspect-cch}
head(task_cch$data())
```

## Subcohort-Only Example

For comparison, let's restrict the analysis to the subcohort only:

{r subcohort-only}

cch_config_subcoh <- list(
  subcohort = "subcoh",
  cch_weight = "Borgan1",   # Ignored when subcohort_only = TRUE
  sampling_fraction = 0.2,  # Ignored
  subcohort_only = TRUE
)

target_info_subcoh <- c(id = "lung_subcoh", time = "time", event = "status", type = "right")
task_subcoh <- CreateMyTaskSurv(
  lung_dt, 
  target_info_subcoh, 
  backend_info_cch, 
  cch = cch_config_subcoh
)
print(task_subcoh)
```

This task:

* Filters to `subcoh == 1` rows.
* Doesn't compute weights (uses `backend_info$weight_col` if provided, but here it's `NULL`).
* Labels task as "Dataset: lung_data; Task type: right; CCH: Subcohort only".

## Multi-State Example

Finally, let's simulate a multi-state scenario by creating a `status_mstate` column with 0 (censored), 1 (death from cancer), and 2 (death from other causes). We'll use ph.ecog to simulate this distinction:

```{r multi-state}
lung_dt_mstate <- copy(lung_dt)
lung_dt_mstate[, status_mstate := status]  # Start with 0/1

# Simulate competing event: if dead (1) and ph.ecog >= 2, assume other cause (2)
lung_dt_mstate[status == 1 & ph.ecog >= 2, status_mstate := 2]
lung_dt_mstate[, status_mstate := factor(status_mstate, levels = c("0", "1", "2"))]

target_info_mstate <- c(id = "lung_mstate", time = "time", event = "status_mstate", type = "mstate")
backend_info_mstate <- list(
  id = "lung_data",
  primary_key = "id",
  feature_cols = c("age", "sex", "ph.ecog")
)
cch_config_mstate <- list(
  subcohort = "subcoh",
  cch_weight = "Borgan2",
  sampling_fraction = 0.2,
  subcohort_only = FALSE
)

task_mstate <- CreateMyTaskSurv(
  lung_dt_mstate, 
  target_info_mstate, 
  backend_info_mstate, 
  cch = cch_config_mstate
)
print(task_mstate)
```

This task:

* Uses type = "mstate" with status_mstate as a factor (0 = censored, 1 = cancer death, 2 = other death).
* Applies Borgan2 weights: all subcohort members get 1/0.2 = 5, non-subcohort events get 1.
* Adds ph.ecog as a feature.
 * Labels task as "Dataset: lung_data; Task type: mstate; CCH weights: Borgan2; Sampling fraction: 0.2".

Inspect the multi-state data:

```{r inspect-mstate}
head(task_mstate$data())
```

## Conclusion

The `CreateMyTaskSurv` function provides a flexible way to create `mlr3` survival tasks from a given dataset. 
Starting with a minimal right-censored example, we expanded to case-cohort designs (full cohort and subcohort-only)
and a multi-state scenario. Each task can be used with mlr3 learners for survival analysis, leveraging weights 
and stratification as needed.

For further exploration, try:
Adding more features (e.g., ph.karno, wt.loss).

Adjusting `sampling_fraction` or `cch_weight` methods.

Applying filters (e.g., backend_info$filter = "age < 70").

